---
title: "Introduction to R"
date: "`r Sys.Date()`"
output: 
  html_document: 
    highlight: haddock
    number_sections: yes
    theme: flatly
    toc: yes
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = ">")
options(dplyr.print_min = 4L, dplyr.print_max = 4L)
library(dplyr)
library(ggplot2)
```

# The example data

To explore the basic data manipulation in R, we'll use some fishy data. The dataset contains measurments of 123252 fish of a particular species.

To do: Describe in more detail ...

```{r, echo = FALSE, eval = FALSE}
fish <- read.csv2("species/species-1.csv") %>%
  tbl_df() %>%
  select(-X)
library(readr)
write_csv(fish, "/net/www/export/home/hafri/einarhj/public_html/data/fish.csv")
```

```{r}
fish <- read.csv("http://www.hafro.is/~einarhj/data/fish.csv") %>%
  tbl_df()
dim(fish)
glimpse(fish)
```


# Basic dataframe manipulations

In the following tutorial we are using functions in the `dplyr`-package. As is stated in the [Introduction to dplyr](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html):

> When working with data you must:
>
>* Figure out what you want to do.
>
>* Describe those tasks in the form of a computer program.
>
>* Execute the program.
>
>The dplyr package makes these steps fast and easy:
>
>* By constraining your options, it simplifies how you can think about common data manipulation tasks.
>
>* It provides simple "verbs", functions that correspond to the most common data manipulation tasks, to help you translate those thoughts into code.
>
>* It uses efficient data storage backends, so you spend less time waiting for the computer.
>
> dplyr can work with data frames as is, but if you're dealing with large data, it's worthwhile to convert them to a `tbl_df` (as done above): this is a wrapper around a data frame that won't accidentally print a lot of data to the screen.
>
>Dplyr aims to provide a function for each basic verb of data manipulation:
>
>* `filter()` (and `slice()`)
>* `arrange()`
>* `select()` (and `rename()`)
>* `distinct()`
>* `mutate()` (and `transmute()`)
>* `summarise()`
>* `sample_n()` and `sample_frac()`

The following structure of the text is inspired by the above cited dplyr-tutorial. Any sentences that are taken directly from that text are quoted.

## Select columns with `select()`

"Often you work with large datasets with many columns but only a few are actually of interest to you. `select()` allows you to rapidly zoom in on a useful subset."

Select columns by name:
```{r}
select(fish, year, length, sex)
```

Select all columns between length and ungutted weight (inclusive)
```{r}
select(fish, sex:ungutted.weight)
```

Select all columns except those from sex to species (inclusive)
```{r}
select(fish, -(sex:species))
```


<div class="panel panel-warning">
<div class="panel-heading">A note on piping</div>
<div class="panel-body">

In the `select` function the first argument is the dataframe you want to manipulate (here the fish dataframe). Using a piping function one could have written the above as:

```{r, eval = FALSE}
fish %>% select(year, length, sex)
fish %>% select(sex:ungutted.weight)
fish %>% select(-(sex:species))
```

</div>
</div>



The `select` function works similarly to the `select` argument in `base::subset()`.

"There are a number of helper functions you can use within `select()`, like `starts_with()`, `ends_with()`, `matches()` and `contains()`. These let you quickly match larger blocks of variables that meet some criterion. See `?select` for more details."

## Rename columns with `rename()`

Renaming columns
```{r, eval=FALSE}
rename(fish,
       unguttedWeightInGrams = ungutted.weight,
       )
```

You can rename variables with `select()` by using named arguments:

```{r}
select(fish, uw = ungutted.weight)
```

But because `select()` drops all the variables not explicitly mentioned, it's not that useful. Instead, use `rename()`:

```{r, eval=FALSE}
rename(fish, uw = ungutted.weight)
```

## Filter rows with `filter()`

`filter()` allows you to select a subset of rows in a data frame. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame:

For example we can select all data from year 2003 with:

```{r}
filter(fish, year == 2003)
```

This is equivalent to:
```{r, eval=FALSE}
filter(fish, year %in% c(2003))
```

If we want to select for more years we could do (here both year 2002 and 2004:
```{r, eval=FALSE}
filter(fish, year == 2002 | year == 2003)
```

The `|` stands for __or__. An alternative could be:
```{r, eval=FALSE}
filter(fish, year %in% c(2002:2003))
```

Another example, we can select all female fish from 2003 with:
```{r}
filter(fish, year == 2003, sex == "F")
```

Note that this is equivalent to:
```{r, eval = FALSE}
fish %>% filter(year == 2003, sex == "F")
```

In the latter case the `fish` dataframe is passed/piped into the function `filter`

Both examples above are equivalent to the more verbose code in base R:

```{r, eval = FALSE}
fish[fish$year == 2003 & fish$sex == "F", ]
```

Note that in the above examples filtered data are __not__ stored as object in your environment. For that one needs to do:
```{r}
females_2003 <- filter(fish, year == 2003, sex == "F")
```

which is the same as:
```{r}
females_2003 <- fish %>% filter(year == 2003, sex == "F")
```

Now we have an smaller dataframe with only 1976 rows/observations (the fish dataframe has 123252 rows):
```{r}
dim(females_2003)
```

`filter()` works similarly to `subset()` except that you can give it any number of filtering conditions, which are joined together with `&` (not `&&` which is easy to do accidentally!).  You can also use other boolean operators:

```{r}
x <- filter(fish, year == 2003 | sex == "F")
```

The `|` stands for __or__, so in the above we select all data for year 2003 and only data for females for the other years. We can visualize this by using the `table` function:
```{r}
table(x$year, x$sex)
```

To select rows by position, use `slice()`:

```{r}
slice(fish, 1001:1002)
```
